When Do You Repeat Yourself?\\ Voices from the Trenches of Linux Kernel Maintainers on Code Duplication

### 2025-10-28 papers/techdebt2026/techdebt.tex
techdebt2025: update discussion regarding Takeaway 3
Signed-off-by: Paulo Meirelles <paulormm@ime.usp.br>

Since we suffered a delay in the review process, we directly contacted a maintainer to understand why. From this conversation, we understand that one reason for the delay was that duplicated code enhances the independence of GPU driver code. This approach allows developers to make changes to a specific GPU without needing to test compatibility with others, which helps save a significant amount of time and effort. This example of hardware variations, also mentioned by Kapser and Godfrey~\cite{cloneharm}, leads us to the primary conclusion of our complete participant observation:

\begin{highlight-box}{Takeaway T1}
Driver developers do not necessarily view duplicated code negatively, which contradicts standard software engineering practices.
\end{highlight-box}

@@

Regarding our analysis from the \textbf{participant-as-observer} phase, we identified that most groups applied the parameterized refactor method, while a few used the extractor method. For instance, Groups 1 and 4 worked on similar duplications, where two functions were involved: one to determine if a register was volatile and another to determine if it was writable. These two functions returned opposite Boolean results. Group 1 initially addressed the problem using the extract method, while Groups 2 and 4 used macros. After feedback from maintainers, both solutions converged toward keeping one of the functions while modifying the other to return a negated query of the former function.

By the end of the feedback cycle, the patch from Group 4 was accepted, but the final implementation differed substantially from the original. The change was primarily driven by a maintainer's feedback: \textit{``I think the old code is more readable than hiding the values in a macro even if it is duplicating a few lines of code.''}
%
This response category revealed a recurring concern among maintainers that went beyond functional correctness or redundancy reduction. Instead, it reflected a preference for code clarity and comprehension. Building on these consistent insights from multiple review interactions, we formulated our second takeaway:

\begin{highlight-box}{Takeaway T2 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the \textbf{DRY principle} was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

Maintainers considered the patch from Group 1 irrelevant.
The reason is that Group 1 chose to apply the refactoring to a device that contained a register that was both read-only and non-volatile. Thus, the technical debt selected by Group 1 was not merely a duplication but an incorrect implementation of the method.

\noindent
\begin{footnotesize}
\textit{
  Quote: ``[...] such patches might not worth it since the proposed
  improvement is very small (and questionable) while the upstreaming process
  still requires some effort.''
}
\end{footnotesize}

\begin{highlight-box}{Takeaway T3 (Integration Overhead)}
  A maintainer suggested that the effort put into the contribution after it was
  accepted (i.e., the process of propagating the change to upper maintainers)
  was considered. In other words, maintainers may \textbf{evaluate the
  trade-off between the impact of contributions (not only deduplications) and
  the struggle to integrate them into the complex Linux project}.
\end{highlight-box}

Groups 3, 6, and 10 created a struct to encapsulate the parameters and remove the duplication by applying the parameterized method. The patches from Groups 6 and 10 were rejected by the same maintainer who approved the patch from Group 3, tackling the duplication of a simple iterative algorithm, which also contained a bug that could be inferred by inspecting both functions. In this case, the group refactored code from a more critical technical debt and took the opportunity to fix an unnoticed incorrect behavior.

The maintainer's justification for rejecting the patch from Group 6 was that it proposed to fix a duplication while adding more lines than removing, and that it was not convincing enough in other aspects, such as readability.

\begin{highlight-box}{Takeaway T4 (Cognitive Load)}
  Needing to jump to a function definition (to see its behavior) or a function
  call (to see its concrete parameters) in order to understand code is a
  cognitively demanding task for developers~\cite{questions-programmers-ask}.
  Duplications that avoided this were justifiable to reduce cognitive load when
  parsing through the code.
\hfill\\

\noindent
\begin{footnotesize}
\textit{
  Quote: ``Wrapping this up doesn't provide any real advantage, requiring as
  it does the reviewer to look at this function AND where the value is set
  rather than seeing them in one place.''
}
\end{footnotesize}
\end{highlight-box}

@@

%TODO ADICIONADO APENAS PARA FECHAR 5 PGS CERTINHO. ME REMOVA!!!
\hfill\\

>>>>

Since we suffered a delay in the review process, we directly contacted a maintainer to understand why. From this conversation, we understand that one reason for the delay was that duplicated code enhances the independence of GPU driver code. This approach allows developers to make changes to a specific GPU without needing to test compatibility with others, which helps save a significant amount of time and effort. This example of hardware variations, also mentioned by Kapser and Godfrey~\cite{cloneharm}, leads us to the primary conclusion of our complete participant observation:
%
\begin{highlight-box}{Takeaway T1}
Driver developers do not necessarily view duplicated code negatively, which contradicts standard software engineering practices.
\end{highlight-box}

@@

Regarding our analysis from the \textbf{participant-as-observer} phase, we identified that most groups applied the parameterized refactor method, while a few used the extractor method. For instance, Groups 1 and 4 worked on similar duplications, where two functions were involved: one to determine if a register was volatile and another to determine if it was writable. These two functions returned opposite Boolean results. Group 1 initially addressed the problem using the extract method, while Groups 2 and 4 used macros. After feedback from maintainers, all solutions converged toward keeping one of the functions while modifying the other to return a negated query of the former function.

By the end of the feedback cycle, the patch from Group 4 was accepted, but the final implementation differed substantially from the original without using macros. The change was primarily driven by a maintainer's feedback: \textit{``I think the old code is more readable than hiding the values in a macro even if it is duplicating a few lines of code.''}.
%
This response category (also from groups 1, 2, and 10) revealed a recurring concern among maintainers that went beyond functional correctness or redundancy reduction. Instead, it reflected a preference for code clarity and comprehension. Building on these consistent insights from multiple review interactions, we formulated our second takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T2 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the \textbf{DRY principle} was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

Still focusing on Group 1, maintainers considered the patch irrelevant. The reason was that Group 1 chose to apply the refactoring to a device containing a register that was both read-only and non-volatile. Thus, the technical debt selected by Group 1 was not merely a duplication but an incorrect implementation of the method.
%
While the previous interactions highlighted how maintainers often favor readability over strict code deduplication, this case added a new dimension to our understanding: the \textit{cost-benefit} evaluation of proposed changes. A maintainer pointed out that the effort required to propagate such modifications after acceptance (i.e., during the upstreaming process) could outweigh the perceived benefits: ``\textit{[...] such patches might not worth it since the proposed improvement is very small (and questionable) while the upstreaming process still requires some effort.}''
From this feedback, we crafted our third takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T3 (Integration Overhead)}
Maintainers may evaluate the trade-off between the impact of contributions
(not only deduplications) and the struggle to integrate them into the complex
Linux project.
\end{highlight-box}

%TODO: David, consolidar a partir daqui
Groups 3, 6, and 10 created a struct to encapsulate the parameters and remove the duplication by applying the parameterized method. The patches from Groups 6 and 10 were rejected by the same maintainer who approved the patch from Group 3, tackling the duplication of a simple iterative algorithm, which also contained a bug that could be inferred by inspecting both functions. In this case, the group refactored code from a more critical technical debt and took the opportunity to fix an unnoticed incorrect behavior.

The maintainer's justification for rejecting the patch from Group 6 was that it proposed to fix a duplication while adding more lines than removing, and that it was not convincing enough in other aspects, such as readability. ... ``Wrapping this up doesn't provide any real advantage, requiring as
  it does the reviewer to look at this function AND where the value is set
  rather than seeing them in one place.''

\begin{highlight-box}{Takeaway T4 (Cognitive Load)}
  Needing to jump to a function definition (to see its behavior) or a function
  call (to see its concrete parameters) in order to understand code is a
  cognitively demanding task for developers~\cite{questions-programmers-ask}.
  Duplications that avoided this were justifiable to reduce cognitive load when
  parsing through the code.
\end{highlight-box}

@@
